---
title: "ST558 Project 3 EDA"
author: "Liam Flaherty"
format: html
editor: visual
---

## EDA

We are given some data on diabetes from <https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/>. We download the file locally and load it as a tibble into our R session.

```{r}
library(tidyverse)

diabetes=read_csv("diabetes.csv", show_col_types = FALSE)
diabetes
```

We would like to get a preliminary look at the data, which we can do with normal summary functions. 

```{r}
str(diabetes, max.level=1, give.attr=FALSE)
summary(diabetes)
```

All of our data is numeric, and most of our data is binary. Most of this data is easily interpretable (e.g. in the `HighBP` predictor, 1 means the observation has high blood pressure, and 0 means that the observation doesn't have high bloop pressure). A few non-binary predictors deserve explanations. The `GenHlth` variable is the respondents personal rating of their health on a scale of 1 (excellent) to 5 (poor). The `MentHlth` and `PhyscHlth` predictors are the respondents personal recollection for how many of the last thirty days that their mental or physical health was not good. The `Age` variable is categorized into 13 different levels with ages 18-24 coded as 1, and ages 80 and up coded as 13. The `Education` variable is graded on a scale of 1-6, with 1 representing no school past kindergarten, and 6 representing completing 4 years of college. The `Income` variable is graded on a scale of 1-8 with 1 representing earning less than $10,000, and 8 representing earning more than $75,000.
